---
title: "15 Insights from South Dakota School Enrollment Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{15 Insights from South Dakota School Enrollment Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5,
  cache = TRUE
)
```

```{r load-packages}
library(sdschooldata)
library(dplyr)
library(tidyr)
library(ggplot2)

theme_set(theme_minimal(base_size = 14))
```

This vignette explores South Dakota's public school enrollment data, surfacing key trends and demographic patterns across 20 years of data (2006-2025).

---

## 1. South Dakota enrollment peaked in 2022 and is now declining

South Dakota's public school enrollment grew steadily from 2015 to 2022, peaking at 141,429, but has dropped 1.8% since then -- losing 2,568 students in three years.

```{r statewide-trend}
enr <- fetch_enr_multi(c(2015:2020, 2022:2025), use_cache = TRUE)

state_totals <- enr |>
  filter(is_state, subgroup == "total_enrollment", grade_level == "TOTAL") |>
  select(end_year, n_students) |>
  mutate(change = n_students - lag(n_students),
         pct_change = round(change / lag(n_students) * 100, 2))

stopifnot(nrow(state_totals) > 0)
state_totals
```

```{r statewide-chart}
ggplot(state_totals, aes(x = end_year, y = n_students)) +
  geom_vline(xintercept = 2020, linetype = "dashed", color = "gray50", alpha = 0.7) +
  annotate("text", x = 2020, y = max(state_totals$n_students), label = "COVID", hjust = -0.2, color = "gray50", size = 3) +
  geom_line(linewidth = 1.2, color = "#003087") +
  geom_point(size = 3, color = "#003087") +
  scale_y_continuous(labels = scales::comma) +
  scale_x_continuous(breaks = seq(2015, 2025, 2)) +
  labs(
    title = "South Dakota Public School Enrollment (2015-2025)",
    subtitle = "Peaked in 2022 at 141,429; now declining",
    x = "School Year (ending)",
    y = "Total Enrollment"
  )
```

---

## 2. Sioux Falls dominates the state

The Sioux Falls School District is by far the largest in the state, with more students than the next several districts combined. Rapid City is a distant second.

```{r top-districts}
enr_2025 <- fetch_enr(2025, use_cache = TRUE)

top_10 <- enr_2025 |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL") |>
  arrange(desc(n_students)) |>
  head(10) |>
  select(district_name, n_students)

stopifnot(nrow(top_10) > 0)
top_10
```

```{r top-districts-chart}
top_10 |>
  mutate(district_name = forcats::fct_reorder(district_name, n_students)) |>
  ggplot(aes(x = n_students, y = district_name)) +
  geom_col(fill = "#003087") +
  scale_x_continuous(labels = scales::comma) +
  labs(
    title = "South Dakota's 10 Largest School Districts (2025)",
    x = "Total Enrollment",
    y = NULL
  )
```

---

## 3. Native American students are a significant population

South Dakota has one of the highest percentages of Native American students in the nation, reflecting the state's large reservation lands including Pine Ridge, Rosebud, and Standing Rock. Demographic data is reported at the campus level; here we aggregate across all campuses statewide.

```{r demographics}
state_total <- enr_2025 |>
  filter(is_state, subgroup == "total_enrollment", grade_level == "TOTAL") |>
  pull(n_students)

demographics <- enr_2025 |>
  filter(is_campus, grade_level == "TOTAL",
         subgroup %in% c("white", "native_american", "hispanic", "black", "asian", "multiracial")) |>
  group_by(subgroup) |>
  summarize(n_students = sum(n_students, na.rm = TRUE), .groups = "drop") |>
  mutate(pct = round(n_students / state_total * 100, 1)) |>
  arrange(desc(n_students))

stopifnot(nrow(demographics) > 0)
demographics
```

```{r demographics-chart}
demographics |>
  mutate(subgroup = forcats::fct_reorder(subgroup, n_students)) |>
  ggplot(aes(x = n_students, y = subgroup, fill = subgroup)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = paste0(pct, "%")), hjust = -0.1) +
  scale_x_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.15))) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "South Dakota Student Demographics (2025)",
    subtitle = "Aggregated from campus-level enrollment reports",
    x = "Number of Students",
    y = NULL
  )
```

---

## 4. Sioux Falls grows while Rapid City and rural hubs decline

The state's largest district continues to grow, but Rapid City lost 12% of its enrollment since 2015 and other regional hubs are shrinking. District names vary across years in SD data, so we use district_id to track districts consistently.

```{r urban-growth}
urban_growth <- enr |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL",
         grepl("Sioux Falls|Rapid City|Aberdeen|Brookings|Watertown", district_name)) |>
  group_by(district_id) |>
  arrange(end_year) |>
  summarize(
    district_name = last(district_name),
    y2015 = n_students[end_year == min(end_year)],
    y2025 = n_students[end_year == max(end_year)],
    pct_change = round((y2025 / y2015 - 1) * 100, 1),
    .groups = "drop"
  ) |>
  arrange(desc(pct_change))

stopifnot(nrow(urban_growth) > 0)
urban_growth
```

```{r regional-chart}
enr |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL",
         grepl("Sioux Falls|Rapid City|Aberdeen|Brookings", district_name)) |>
  mutate(district_label = case_when(
    district_id == "49005" ~ "Sioux Falls",
    district_id == "51004" ~ "Rapid City",
    district_id == "06001" ~ "Aberdeen",
    district_id == "05001" ~ "Brookings",
    TRUE ~ district_name
  )) |>
  ggplot(aes(x = end_year, y = n_students, color = district_label)) +
  geom_vline(xintercept = 2020, linetype = "dashed", color = "gray50", alpha = 0.7) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "South Dakota Urban District Growth",
    subtitle = "Sioux Falls grows while Rapid City and rural hubs lose students",
    x = "School Year",
    y = "Enrollment",
    color = "District"
  )
```

---

## 5. Many tiny rural districts

South Dakota has a large number of very small school districts, many with fewer than 200 students, reflecting the state's rural character and sparse population.

```{r small-districts}
small <- enr_2025 |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL") |>
  filter(n_students < 200) |>
  arrange(n_students) |>
  head(15) |>
  select(district_name, n_students)

stopifnot(nrow(small) > 0)
small
```

---

## 6. Hispanic enrollment is rising fast

Hispanic students are the fastest-growing demographic group in South Dakota schools, climbing from 7.97% to 9.25% of statewide enrollment in just four years (2022-2025). Campus-level demographic data is available starting in 2022.

```{r hispanic-trend}
hispanic_years <- c(2022:2025)
enr_hispanic <- fetch_enr_multi(hispanic_years, use_cache = TRUE)

hispanic_trend <- enr_hispanic |>
  filter(is_campus, subgroup == "hispanic", grade_level == "TOTAL") |>
  group_by(end_year) |>
  summarize(n_students = sum(n_students, na.rm = TRUE), .groups = "drop")

state_totals_hisp <- enr_hispanic |>
  filter(is_state, subgroup == "total_enrollment", grade_level == "TOTAL") |>
  select(end_year, total = n_students)

hispanic_trend <- hispanic_trend |>
  left_join(state_totals_hisp, by = "end_year") |>
  mutate(pct = round(n_students / total * 100, 2)) |>
  select(end_year, n_students, pct)

stopifnot(nrow(hispanic_trend) > 0)
hispanic_trend
```

```{r growth-chart}
ggplot(hispanic_trend, aes(x = end_year, y = n_students)) +
  geom_line(linewidth = 1.2, color = "#2E8B57") +
  geom_point(size = 3, color = "#2E8B57") +
  geom_text(aes(label = paste0(pct, "%")), vjust = -1.5) +
  scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = "Hispanic Student Enrollment Growth (2022-2025)",
    subtitle = "Fastest-growing demographic group in South Dakota",
    x = "School Year",
    y = "Number of Students"
  )
```

---

## 7. Sioux Falls suburban growth outpaces the city

Harrisburg and Tea Area have seen explosive growth as Sioux Falls suburbs boom, with Harrisburg more than doubling its enrollment from 2,724 to 6,398 students in just 15 years.

```{r suburban-growth}
suburbs <- fetch_enr_multi(c(2011, 2015, 2020, 2025), use_cache = TRUE)

suburb_trend <- suburbs |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL",
         grepl("Harrisburg|Tea Area|Brandon Valley", district_name)) |>
  select(end_year, district_name, n_students)

stopifnot(nrow(suburb_trend) > 0)
suburb_trend
```

```{r suburban-growth-chart}
suburbs |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL",
         grepl("Harrisburg|Tea Area|Brandon Valley", district_name)) |>
  ggplot(aes(x = end_year, y = n_students, color = district_name)) +
  geom_vline(xintercept = 2020, linetype = "dashed", color = "gray50", alpha = 0.7) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_y_continuous(labels = scales::comma) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "Sioux Falls Suburban District Growth",
    subtitle = "Harrisburg, Tea Area, and Brandon Valley boom",
    x = "School Year",
    y = "Total Enrollment",
    color = "District"
  )
```

---

## 8. Rapid City: West River anchor

Rapid City Area School District anchors western South Dakota, serving as the only major urban district west of the Missouri River.

```{r rapid-city}
rapid <- fetch_enr_multi(c(2015:2020, 2022:2025), use_cache = TRUE)

rapid_trend <- rapid |>
  filter(is_district, grepl("Rapid City", district_name),
         subgroup == "total_enrollment", grade_level == "TOTAL") |>
  select(end_year, n_students)

stopifnot(nrow(rapid_trend) > 0)
rapid_trend
```

```{r rapid-city-chart}
ggplot(rapid_trend, aes(x = end_year, y = n_students)) +
  geom_vline(xintercept = 2020, linetype = "dashed", color = "gray50", alpha = 0.7) +
  annotate("text", x = 2020, y = 15500, label = "COVID", hjust = -0.2, color = "gray50", size = 3) +
  geom_line(linewidth = 1.2, color = "#8B4513") +
  geom_point(size = 3, color = "#8B4513") +
  scale_y_continuous(labels = scales::comma, limits = c(12000, 16000)) +
  labs(
    title = "Rapid City Area School District Enrollment",
    subtitle = "Steady decline from 13,832 peak in 2018 to 12,040 in 2025",
    x = "School Year",
    y = "Total Enrollment"
  )
```

---

## 9. Reservation schools: Todd County and Pine Ridge

Districts serving reservation communities face unique challenges. Todd County (Rosebud) and Oglala Lakota County (Pine Ridge) serve predominantly Native American students.

```{r reservation-schools}
reservation <- fetch_enr_multi(c(2015, 2020, 2025), use_cache = TRUE)

res_data <- reservation |>
  filter(is_district,
         grepl("Todd County|Oglala Lakota|Shannon", district_name),
         subgroup == "total_enrollment",
         grade_level == "TOTAL")

stopifnot(nrow(res_data) > 0)
res_data
```

```{r reservation-chart}
reservation |>
  filter(is_district,
         grepl("Todd County|Oglala Lakota", district_name),
         subgroup == "total_enrollment",
         grade_level == "TOTAL") |>
  ggplot(aes(x = end_year, y = n_students, fill = district_name)) +
  geom_col(position = "dodge") +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Reservation School District Enrollment",
    subtitle = "Todd County (Rosebud) and Oglala Lakota County (Pine Ridge)",
    x = "School Year",
    y = "Enrollment",
    fill = "District"
  )
```

---

## 10. Rural consolidation pressure

Many of South Dakota's smallest districts face consolidation pressure. Districts with fewer than 100 students struggle with economies of scale.

```{r rural-tiny}
tiny <- fetch_enr(2025, use_cache = TRUE)

tiny_districts <- tiny |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL",
         n_students < 100) |>
  arrange(n_students) |>
  head(20) |>
  select(district_name, n_students)

stopifnot(nrow(tiny_districts) > 0)
tiny_districts
```

```{r rural-distribution-chart}
size_dist <- fetch_enr(2025, use_cache = TRUE) |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL") |>
  mutate(size_category = cut(n_students,
                              breaks = c(0, 100, 250, 500, 1000, 5000, Inf),
                              labels = c("<100", "100-249", "250-499", "500-999", "1000-4999", "5000+"))) |>
  count(size_category)

stopifnot(nrow(size_dist) > 0)

ggplot(size_dist, aes(x = size_category, y = n, fill = size_category)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = n), vjust = -0.5) +
  scale_fill_viridis_d(option = "C") +
  labs(
    title = "South Dakota Districts by Enrollment Size",
    subtitle = "Many tiny districts face consolidation pressure",
    x = "District Size",
    y = "Number of Districts"
  )
```

---

## 11. High school enrollment surging while elementary stays flat

South Dakota's high school enrollment grew 12% from 2015 to 2025, while K-8 enrollment stayed essentially flat. This reflects larger birth cohorts aging into upper grades.

```{r hs-vs-elem}
hs_elem <- enr |>
  filter(is_state, subgroup == "total_enrollment",
         grade_level %in% c("K", paste0("0", 1:8), "09", "10", "11", "12")) |>
  mutate(level = ifelse(grade_level %in% c("09", "10", "11", "12"), "High School", "K-8")) |>
  group_by(end_year, level) |>
  summarize(n_students = sum(n_students, na.rm = TRUE), .groups = "drop")

stopifnot(nrow(hs_elem) > 0)
hs_elem
```

```{r hs-vs-elem-chart}
ggplot(hs_elem, aes(x = end_year, y = n_students, color = level)) +
  geom_vline(xintercept = 2020, linetype = "dashed", color = "gray50", alpha = 0.7) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  scale_y_continuous(labels = scales::comma) +
  scale_color_manual(values = c("High School" = "#E7298A", "K-8" = "#1B9E77")) +
  labs(
    title = "K-8 vs High School Enrollment (2015-2025)",
    subtitle = "High school enrollment grew 12% while K-8 stayed flat",
    x = "School Year",
    y = "Enrollment",
    color = "Level"
  )
```

---

## 12. The Black Hills corridor

The Black Hills region forms a distinct educational corridor, with Rapid City at its center and smaller communities like Spearfish, Custer, and Belle Fourche serving surrounding areas.

```{r black-hills}
black_hills <- fetch_enr(2025, use_cache = TRUE)

bh_districts <- black_hills |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL",
         grepl("Rapid City|Spearfish|Sturgis|Custer|Lead|Deadwood|Belle Fourche", district_name)) |>
  arrange(desc(n_students)) |>
  select(district_name, n_students)

stopifnot(nrow(bh_districts) > 0)
bh_districts
```

```{r black-hills-chart}
fetch_enr(2025, use_cache = TRUE) |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL",
         grepl("Rapid City|Spearfish|Sturgis|Custer|Lead|Belle Fourche", district_name)) |>
  mutate(district_name = forcats::fct_reorder(district_name, n_students)) |>
  ggplot(aes(x = n_students, y = district_name, fill = district_name)) +
  geom_col(show.legend = FALSE) +
  scale_x_continuous(labels = scales::comma) +
  scale_fill_viridis_d(option = "D") +
  labs(
    title = "Black Hills Corridor School Districts",
    subtitle = "Rapid City dominates western South Dakota",
    x = "Total Enrollment",
    y = NULL
  )
```

---

## 13. Aberdeen and the northeast

Aberdeen School District anchors the northeast, with surrounding agricultural communities feeding into regional schools.

```{r northeast}
northeast <- fetch_enr_multi(c(2015:2020, 2022:2025), use_cache = TRUE)

ne_trend <- northeast |>
  filter(is_district, grepl("Aberdeen", district_name),
         subgroup == "total_enrollment", grade_level == "TOTAL") |>
  select(end_year, n_students)

stopifnot(nrow(ne_trend) > 0)
ne_trend
```

```{r northeast-chart}
fetch_enr_multi(c(2015:2020, 2022:2025), use_cache = TRUE) |>
  filter(is_district,
         grepl("Aberdeen|Watertown|Huron|Mitchell", district_name),
         subgroup == "total_enrollment", grade_level == "TOTAL") |>
  ggplot(aes(x = end_year, y = n_students, color = district_name)) +
  geom_vline(xintercept = 2020, linetype = "dashed", color = "gray50", alpha = 0.7) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  scale_y_continuous(labels = scales::comma) +
  scale_color_brewer(palette = "Dark2") +
  labs(
    title = "Regional Hub Districts: East River",
    subtitle = "Aberdeen, Watertown, Huron, and Mitchell anchor their regions",
    x = "School Year",
    y = "Enrollment",
    color = "District"
  )
```

---

## 14. Pre-K enrollment patterns

Pre-kindergarten enrollment varies significantly across districts, reflecting different local policies and access to early childhood programs.

```{r prek}
prek <- fetch_enr(2025, use_cache = TRUE)

prek_data <- prek |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "PK") |>
  filter(n_students > 0) |>
  arrange(desc(n_students)) |>
  head(15) |>
  select(district_name, n_students)

stopifnot(nrow(prek_data) > 0)
prek_data
```

```{r prek-chart}
prek_chart_data <- fetch_enr(2025, use_cache = TRUE) |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "PK") |>
  filter(n_students > 0) |>
  arrange(desc(n_students)) |>
  head(10) |>
  mutate(district_name = forcats::fct_reorder(district_name, n_students))

stopifnot(nrow(prek_chart_data) > 0)

ggplot(prek_chart_data, aes(x = n_students, y = district_name)) +
  geom_col(fill = "#9467BD") +
  scale_x_continuous(labels = scales::comma) +
  labs(
    title = "Top 10 Districts by Pre-K Enrollment",
    subtitle = "Urban districts lead in early childhood programs",
    x = "Pre-K Students",
    y = NULL
  )
```

---

## 15. Multiracial students: South Dakota's growing diversity

Multiracial students grew from 8,129 (5.75%) to 8,681 (6.25%) of statewide enrollment between 2022 and 2025, reflecting changing family patterns statewide. Campus-level demographic data is available starting in 2022.

```{r multiracial}
multi <- fetch_enr_multi(c(2022, 2023, 2024, 2025), use_cache = TRUE)

multi_trend <- multi |>
  filter(is_campus, subgroup == "multiracial", grade_level == "TOTAL") |>
  group_by(end_year) |>
  summarize(n_students = sum(n_students, na.rm = TRUE), .groups = "drop")

state_totals_multi <- multi |>
  filter(is_state, subgroup == "total_enrollment", grade_level == "TOTAL") |>
  select(end_year, total = n_students)

multi_trend <- multi_trend |>
  left_join(state_totals_multi, by = "end_year") |>
  mutate(pct = round(n_students / total * 100, 2)) |>
  select(end_year, n_students, pct)

stopifnot(nrow(multi_trend) > 0)
multi_trend
```

```{r multiracial-chart}
ggplot(multi_trend, aes(x = end_year, y = n_students)) +
  geom_col(fill = "#17BECF", width = 0.7) +
  geom_text(aes(label = scales::comma(n_students)), vjust = -0.5) +
  scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = "Multiracial Student Enrollment Growth (2022-2025)",
    subtitle = "Growing share of South Dakota's student population",
    x = "School Year",
    y = "Number of Students"
  )
```

---

## Summary

South Dakota's school enrollment data reveals:

- **Post-peak decline**: Enrollment peaked at 141,429 in 2022 and is now dropping
- **Urban concentration**: Sioux Falls and Rapid City dominate enrollment
- **Native American presence**: Significant Native American student population
- **Suburban boom**: Harrisburg and Tea Area growing rapidly
- **Reservation challenges**: Todd and Oglala Lakota counties serve unique populations
- **Rural struggles**: Many tiny districts face consolidation pressure
- **Demographic change**: Hispanic and multiracial enrollment growing steadily
- **High school surge**: HS enrollment grew 12% while K-8 stayed flat
- **Regional hubs**: Black Hills and northeast districts anchor their regions
- **Early childhood**: Pre-K access varies significantly across districts

These trends have implications for school funding, facility planning, and educational equity across the Mount Rushmore State.

---

*Data sourced from the South Dakota Department of Education [Fall Census](https://doe.sd.gov/ofm/enrollment.aspx).*

---

## Session Info

```{r session-info}
sessionInfo()
```
