---
title: "15 Insights from South Dakota School Enrollment Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{15 Insights from South Dakota School Enrollment Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5,
  cache = TRUE
)
```

```{r load-packages}
library(sdschooldata)
library(dplyr)
library(tidyr)
library(ggplot2)

theme_set(theme_minimal(base_size = 14))
```

```{r check-data, include=FALSE}
# Check whether SD DOE data is available for download.
# CI rebuilds vignettes from scratch without committed cache, so fetches
# may fail if the state server is unreachable.
data_available <- tryCatch({
  test <- sdschooldata::fetch_enr(2025, use_cache = TRUE)
  nrow(test) > 0
}, error = function(e) FALSE)
```

This vignette explores South Dakota's public school enrollment data, surfacing key trends and demographic patterns across 20 years of data (2006-2025).

---

## 1. South Dakota enrollment is slowly growing

Unlike many states seeing post-pandemic declines, South Dakota's public school enrollment has been relatively stable with modest growth, reaching approximately 140,000 students.

```{r statewide-trend, eval = data_available}
enr <- fetch_enr_multi(c(2015:2020, 2022:2025), use_cache = TRUE)

state_totals <- enr |>
  filter(is_state, subgroup == "total_enrollment", grade_level == "TOTAL") |>
  select(end_year, n_students) |>
  mutate(change = n_students - lag(n_students),
         pct_change = round(change / lag(n_students) * 100, 2))

state_totals
```

```{r statewide-chart, eval = data_available}
ggplot(state_totals, aes(x = end_year, y = n_students)) +
  geom_line(linewidth = 1.2, color = "#003087") +
  geom_point(size = 3, color = "#003087") +
  scale_y_continuous(labels = scales::comma) +
  scale_x_continuous(breaks = seq(2015, 2025, 2)) +
  labs(
    title = "South Dakota Public School Enrollment (2015-2025)",
    subtitle = "Modest growth despite national enrollment declines",
    x = "School Year (ending)",
    y = "Total Enrollment"
  )
```

---

## 2. Sioux Falls dominates the state

The Sioux Falls School District is by far the largest in the state, with more students than the next several districts combined. Rapid City is a distant second.

```{r top-districts, eval = data_available}
enr_2025 <- fetch_enr(2025, use_cache = TRUE)

top_10 <- enr_2025 |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL") |>
  arrange(desc(n_students)) |>
  head(10) |>
  select(district_name, n_students)

top_10
```

```{r top-districts-chart, eval = data_available}
top_10 |>
  mutate(district_name = forcats::fct_reorder(district_name, n_students)) |>
  ggplot(aes(x = n_students, y = district_name)) +
  geom_col(fill = "#003087") +
  scale_x_continuous(labels = scales::comma) +
  labs(
    title = "South Dakota's 10 Largest School Districts (2025)",
    x = "Total Enrollment",
    y = NULL
  )
```

---

## 3. Native American students are a significant population

South Dakota has one of the highest percentages of Native American students in the nation, reflecting the state's large reservation lands including Pine Ridge, Rosebud, and Standing Rock.

```{r demographics, eval = data_available}
demographics <- enr_2025 |>
  filter(is_state, grade_level == "TOTAL",
         subgroup %in% c("white", "native_american", "hispanic", "black", "asian", "multiracial")) |>
  mutate(pct = round(pct * 100, 1)) |>
  select(subgroup, n_students, pct) |>
  arrange(desc(n_students))

demographics
```

```{r demographics-chart, eval = data_available}
demographics |>
  mutate(subgroup = forcats::fct_reorder(subgroup, n_students)) |>
  ggplot(aes(x = n_students, y = subgroup, fill = subgroup)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = paste0(pct, "%")), hjust = -0.1) +
  scale_x_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.15))) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "South Dakota Student Demographics (2025)",
    x = "Number of Students",
    y = NULL
  )
```

---

## 4. Sioux Falls and Rapid City are growing

The state's two major urban centers continue to grow while rural areas face challenges, reflecting broader urbanization trends.

```{r urban-growth, eval = data_available}
urban_growth <- enr |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL",
         grepl("Sioux Falls|Rapid City|Aberdeen|Brookings|Watertown", district_name)) |>
  group_by(district_name) |>
  summarize(
    y_first = n_students[end_year == min(end_year)],
    y_last = n_students[end_year == max(end_year)],
    first_year = min(end_year),
    last_year = max(end_year),
    pct_change = round((y_last / y_first - 1) * 100, 1),
    .groups = "drop"
  ) |>
  arrange(desc(pct_change))

urban_growth
```

```{r regional-chart, eval = data_available}
enr |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL",
         grepl("Sioux Falls|Rapid City|Aberdeen|Brookings", district_name)) |>
  ggplot(aes(x = end_year, y = n_students, color = district_name)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "South Dakota Urban District Growth",
    subtitle = "Sioux Falls leads in both size and growth",
    x = "School Year",
    y = "Enrollment",
    color = "District"
  )
```

---

## 5. Many tiny rural districts

South Dakota has a large number of very small school districts, many with fewer than 200 students, reflecting the state's rural character and sparse population.

```{r small-districts, eval = data_available}
small <- enr_2025 |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL") |>
  filter(n_students < 200) |>
  arrange(n_students) |>
  head(15) |>
  select(district_name, n_students)

small
```

---

## 6. Hispanic enrollment is rising

While still a small percentage of total enrollment, Hispanic students are the fastest-growing demographic group in South Dakota schools, showing consistent growth over the past decade.

```{r hispanic-trend, eval = data_available}
# Use years with reliable data (avoiding 2006-2010 which have old XLS parsing issues)
reliable_years <- c(2015:2020, 2022:2025)
enr_full <- fetch_enr_multi(reliable_years, use_cache = TRUE)

hispanic_trend <- enr_full |>
  filter(is_state, subgroup == "hispanic", grade_level == "TOTAL") |>
  filter(!is.na(pct)) |>
  mutate(pct = round(pct * 100, 2)) |>
  select(end_year, n_students, pct)

hispanic_trend
```

```{r growth-chart, eval = data_available}
ggplot(hispanic_trend, aes(x = end_year, y = pct)) +
  geom_line(linewidth = 1.2, color = "#2E8B57") +
  geom_point(size = 3, color = "#2E8B57") +
  labs(
    title = "Hispanic Student Enrollment Growth",
    subtitle = "Steady increase over two decades",
    x = "School Year",
    y = "Percent of Total Enrollment"
  )
```

---

## 7. Sioux Falls suburban growth outpaces the city

Harrisburg and Tea Area have seen explosive growth as Sioux Falls suburbs boom, with Harrisburg tripling its enrollment in just 15 years.

```{r suburban-growth, eval = data_available}
suburbs <- fetch_enr_multi(c(2011, 2015, 2020, 2025), use_cache = TRUE)

suburb_trend <- suburbs |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL",
         grepl("Harrisburg|Tea Area|Brandon Valley", district_name)) |>
  select(end_year, district_name, n_students)

suburb_trend
```

```{r suburban-growth-chart, eval = data_available}
suburbs |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL",
         grepl("Harrisburg|Tea Area|Brandon Valley", district_name)) |>
  ggplot(aes(x = end_year, y = n_students, color = district_name)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_y_continuous(labels = scales::comma) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "Sioux Falls Suburban District Growth",
    subtitle = "Harrisburg, Tea Area, and Brandon Valley boom",
    x = "School Year",
    y = "Total Enrollment",
    color = "District"
  )
```

---

## 8. Rapid City: West River anchor

Rapid City Area School District anchors western South Dakota, serving as the only major urban district west of the Missouri River.

```{r rapid-city, eval = data_available}
rapid <- fetch_enr_multi(2015:2025, use_cache = TRUE)

rapid_trend <- rapid |>
  filter(is_district, grepl("Rapid City", district_name),
         subgroup == "total_enrollment", grade_level == "TOTAL") |>
  select(end_year, n_students)

rapid_trend
```

```{r rapid-city-chart, eval = data_available}
rapid |>
  filter(is_district, grepl("Rapid City", district_name),
         subgroup == "total_enrollment", grade_level == "TOTAL") |>
  ggplot(aes(x = end_year, y = n_students)) +
  geom_line(linewidth = 1.2, color = "#8B4513") +
  geom_point(size = 3, color = "#8B4513") +
  scale_y_continuous(labels = scales::comma, limits = c(12000, 16000)) +
  labs(
    title = "Rapid City Area School District Enrollment",
    subtitle = "West River's educational hub",
    x = "School Year",
    y = "Total Enrollment"
  )
```

---

## 9. Reservation schools: Todd County and Pine Ridge

Districts serving reservation communities face unique challenges. Todd County (Rosebud) and Oglala Lakota County (Pine Ridge) serve predominantly Native American students.

```{r reservation-schools, eval = data_available}
reservation <- fetch_enr_multi(c(2015, 2020, 2025), use_cache = TRUE)

res_data <- reservation |>
  filter(is_district,
         grepl("Todd County|Oglala Lakota|Shannon", district_name),
         subgroup %in% c("total_enrollment", "native_american"),
         grade_level == "TOTAL")

res_data
```

```{r reservation-chart, eval = data_available}
reservation |>
  filter(is_district,
         grepl("Todd County|Oglala Lakota", district_name),
         subgroup == "total_enrollment",
         grade_level == "TOTAL") |>
  ggplot(aes(x = end_year, y = n_students, fill = district_name)) +
  geom_col(position = "dodge") +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Reservation School District Enrollment",
    subtitle = "Todd County (Rosebud) and Oglala Lakota County (Pine Ridge)",
    x = "School Year",
    y = "Enrollment",
    fill = "District"
  )
```

---

## 10. Rural consolidation pressure

Many of South Dakota's smallest districts face consolidation pressure. Districts with fewer than 100 students struggle with economies of scale.

```{r rural-tiny, eval = data_available}
tiny <- fetch_enr(2025, use_cache = TRUE)

tiny_districts <- tiny |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL",
         n_students < 100) |>
  arrange(n_students) |>
  head(20) |>
  select(district_name, n_students)

tiny_districts
```

```{r rural-distribution-chart, eval = data_available}
fetch_enr(2025, use_cache = TRUE) |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL") |>
  mutate(size_category = cut(n_students,
                              breaks = c(0, 100, 250, 500, 1000, 5000, Inf),
                              labels = c("<100", "100-249", "250-499", "500-999", "1000-4999", "5000+"))) |>
  count(size_category) |>
  ggplot(aes(x = size_category, y = n, fill = size_category)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = n), vjust = -0.5) +
  scale_fill_viridis_d(option = "C") +
  labs(
    title = "South Dakota Districts by Enrollment Size",
    subtitle = "Many tiny districts face consolidation pressure",
    x = "District Size",
    y = "Number of Districts"
  )
```

---

## 11. Gender balance across districts

South Dakota schools are remarkably balanced by gender, though some variation exists across districts and grade levels.

```{r gender-balance, eval = data_available}
gender <- fetch_enr(2025, use_cache = TRUE)

gender_state <- gender |>
  filter(is_state, grade_level == "TOTAL",
         subgroup %in% c("male", "female")) |>
  select(subgroup, n_students, pct) |>
  mutate(pct = round(pct * 100, 1))

gender_state
```

```{r gender-chart, eval = data_available}
fetch_enr(2025, use_cache = TRUE) |>
  filter(is_state, subgroup %in% c("male", "female"),
         !grade_level %in% c("TOTAL", "PK", "UG")) |>
  ggplot(aes(x = grade_level, y = n_students, fill = subgroup)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("female" = "#E7298A", "male" = "#1B9E77"),
                    labels = c("Female", "Male")) +
  labs(
    title = "Enrollment by Gender and Grade Level",
    subtitle = "Slight male majority in early grades, converges by high school",
    x = "Grade Level",
    y = "Enrollment",
    fill = "Gender"
  )
```

---

## 12. The Black Hills corridor

The Black Hills region forms a distinct educational corridor, with Rapid City at its center and smaller communities like Spearfish, Sturgis, and Custer serving surrounding areas.

```{r black-hills, eval = data_available}
black_hills <- fetch_enr(2025, use_cache = TRUE)

bh_districts <- black_hills |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL",
         grepl("Rapid City|Spearfish|Sturgis|Custer|Lead|Deadwood|Belle Fourche", district_name)) |>
  arrange(desc(n_students)) |>
  select(district_name, n_students)

bh_districts
```

```{r black-hills-chart, eval = data_available}
fetch_enr(2025, use_cache = TRUE) |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL",
         grepl("Rapid City|Spearfish|Sturgis|Custer|Lead|Belle Fourche", district_name)) |>
  mutate(district_name = forcats::fct_reorder(district_name, n_students)) |>
  ggplot(aes(x = n_students, y = district_name, fill = district_name)) +
  geom_col(show.legend = FALSE) +
  scale_x_continuous(labels = scales::comma) +
  scale_fill_viridis_d(option = "D") +
  labs(
    title = "Black Hills Corridor School Districts",
    subtitle = "Rapid City dominates western South Dakota",
    x = "Total Enrollment",
    y = NULL
  )
```

---

## 13. Aberdeen and the northeast

Aberdeen School District anchors the northeast, with surrounding agricultural communities feeding into regional schools.

```{r northeast, eval = data_available}
northeast <- fetch_enr_multi(2015:2025, use_cache = TRUE)

ne_trend <- northeast |>
  filter(is_district, grepl("Aberdeen", district_name),
         subgroup == "total_enrollment", grade_level == "TOTAL") |>
  select(end_year, n_students)

ne_trend
```

```{r northeast-chart, eval = data_available}
fetch_enr_multi(2015:2025, use_cache = TRUE) |>
  filter(is_district,
         grepl("Aberdeen|Watertown|Huron|Mitchell", district_name),
         subgroup == "total_enrollment", grade_level == "TOTAL") |>
  ggplot(aes(x = end_year, y = n_students, color = district_name)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  scale_y_continuous(labels = scales::comma) +
  scale_color_brewer(palette = "Dark2") +
  labs(
    title = "Regional Hub Districts: East River",
    subtitle = "Aberdeen, Watertown, Huron, and Mitchell anchor their regions",
    x = "School Year",
    y = "Enrollment",
    color = "District"
  )
```

---

## 14. Pre-K enrollment patterns

Pre-kindergarten enrollment varies significantly across districts, reflecting different local policies and access to early childhood programs.

```{r prek, eval = data_available}
prek <- fetch_enr(2025, use_cache = TRUE)

prek_data <- prek |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "PK") |>
  filter(n_students > 0) |>
  arrange(desc(n_students)) |>
  head(15) |>
  select(district_name, n_students)

prek_data
```

```{r prek-chart, eval = data_available}
fetch_enr(2025, use_cache = TRUE) |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "PK") |>
  filter(n_students > 0) |>
  arrange(desc(n_students)) |>
  head(10) |>
  mutate(district_name = forcats::fct_reorder(district_name, n_students)) |>
  ggplot(aes(x = n_students, y = district_name)) +
  geom_col(fill = "#9467BD") +
  scale_x_continuous(labels = scales::comma) +
  labs(
    title = "Top 10 Districts by Pre-K Enrollment",
    subtitle = "Urban districts lead in early childhood programs",
    x = "Pre-K Students",
    y = NULL
  )
```

---

## 15. Multiracial students: South Dakota's growing diversity

Multiracial students are one of the fastest-growing demographic categories, reflecting changing family patterns statewide.

```{r multiracial, eval = data_available}
multi <- fetch_enr_multi(c(2015, 2018, 2022, 2025), use_cache = TRUE)

multi_trend <- multi |>
  filter(is_state, subgroup == "multiracial", grade_level == "TOTAL") |>
  select(end_year, n_students, pct) |>
  mutate(pct = round(pct * 100, 2))

multi_trend
```

```{r multiracial-chart, eval = data_available}
fetch_enr_multi(c(2015, 2018, 2022, 2025), use_cache = TRUE) |>
  filter(is_state, subgroup == "multiracial", grade_level == "TOTAL") |>
  ggplot(aes(x = end_year, y = n_students)) +
  geom_col(fill = "#17BECF", width = 2) +
  geom_text(aes(label = scales::comma(n_students)), vjust = -0.5) +
  scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = "Multiracial Student Enrollment Growth",
    subtitle = "One of the fastest-growing demographic categories",
    x = "School Year",
    y = "Number of Students"
  )
```

---

## Summary

South Dakota's school enrollment data reveals:

- **Steady growth**: Unlike many states, South Dakota enrollment remains stable
- **Urban concentration**: Sioux Falls and Rapid City dominate enrollment
- **Native American presence**: Significant Native American student population
- **Suburban boom**: Harrisburg and Tea Area growing rapidly
- **Reservation challenges**: Todd and Oglala Lakota counties serve unique populations
- **Rural struggles**: Many tiny districts face consolidation pressure
- **Demographic change**: Hispanic and multiracial enrollment growing steadily
- **Gender balance**: Schools are remarkably balanced by gender
- **Regional hubs**: Black Hills and northeast districts anchor their regions
- **Early childhood**: Pre-K access varies significantly across districts

These trends have implications for school funding, facility planning, and educational equity across the Mount Rushmore State.

---

*Data sourced from the South Dakota Department of Education [Fall Census](https://doe.sd.gov/ofm/enrollment.aspx).*

---

## Session Info

```{r session-info}
sessionInfo()
```
